---
title: "Age filtering"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
path2uisdata <- '~/UNESCO/GEM Report - Data Repository/UIS2021/'
```

## Motivation

Many characteristics, such as whether a given level of schooling has been completed, are of potential interest for *all* individuals in household survey data, but only specific age groups should enter the calculation of defined indicators such as the completion rate.

Since the age restriction is part of the indicator calculation, not a property of the observations, the most general approach is to keep the characteristic without age restriction in the microdata, and only restrict as part of the indicator summary calculation. 

In addition, the good software engineering principle of separating logic from data 

A practical implementation of an approach that meets both goals, i.e. ‘just in time’ age restriction and separation of age range definition (logic) from the specific country-year specific age parameters (data) is outlined in this document.

## The goal

Working backwards, suppose we had a table `indicator_age_scope` containing the ages that are in scope for a given indicator in a given country and year, structured as follows:

```
 iso3c  year indicator     age
 ABC    2010 eduout_prim     6
 ABC    2010 eduout_prim     7
 ABC    2010 eduout_prim     8
 ABC    2010 eduout_prim     9
 ABC    2010 eduout_lsec    10
 ABC    2010 eduout_lsec    11
 ...    ...  ...            ...
```

The age restriction could then be applied with a single line of code for all indicators at once:

```{r, eval=FALSE, include=TRUE}
wide_summarydata <- 
  wide_microdata_long %>% 
  semi_join(indicator_age_scope, by = c('iso3c', 'year', 'indicator', 'age')) %>% 
  wide_summarize()

```

If for a specific analysis we did want an indicator variable in the microdata that indicates who is in scope, this can be added interactively once a trivial helper function is defined:

```{r, eval=FALSE, include=TRUE}

add_age_scope <- function(df, scope) {
  left_join(df, mutate(scope, inscope = TRUE), 
            by = c('iso3c', 'year', 'indicator', 'age')) %>% 
  mutate(inscope = replace_na(inscope, FALSE))
}

wide_microdata_scoped <- add_age_scope(wide_microdata_long, indicator_age_scope)

```

## Implementation

A first step is to **import the relevant parameters for each country-year dyad** from the UIS bulk data download file. A minimal non-redundant set consists of primary entry age plus duration of primary, lower secondary, and upper secondary respectively. From this, all necessary age ranges (e.g. primary school age, 3 to 5 years above final grade of lower secondary, etc.) can be derived.

```{r, echo=TRUE, include=FALSE}
# fetch minimum school structure data from UIS bulk data download file
# not computationally efficient, but concise and transparent
school_structure <- 
  vroom::vroom(paste0(path2uisdata, 'NATMON/NATMON_DATA_NATIONAL.csv'), na = '') %>%
  pivot_wider(names_from = 'INDICATOR_ID', values_from = 'VALUE') %>% 
  select(
    iso3c = COUNTRY_ID, year = YEAR,
    prim_entryage = `299905`, 
    prim_duration = `299932`, 
    lsec_duration = `999976`, 
    usec_duration = `999978`) %>% 
  # to do: extrapolate to bring coverage to all years for all countries
  na.omit %>% 
  filter(iso3c != 'VAT') %>% # exclude Vatican, which has zero school durations
  identity  
```

```{r, echo=TRUE, include=TRUE}
print(school_structure)
```

We define a generic helper function that turns the above structural information into the target structure for `indicator_age_scope`, using the function with the age range definition for a given indicator:

```{r, echo=TRUE}
expand_ages <- function(df, expansion_fun) {
  df %>%
  group_by(iso3c, year) %>% 
  group_modify(~ expansion_fun(.x$prim_entryage, .x$prim_duration, .x$lsec_duration, .x$usec_duration)) %>%
  ungroup
}
```

Given such functions defining indicator-specific age ranges, `indicator_age_scope` can be built by applying them in turn and stacking the results:

```{r, echo=TRUE}
expand_attendance_ages <- function(p0, pd, lsd, usd) {
  tibble(
    indicator = paste('eduout', rep.int(c('prim', 'lsec', 'usec'), c(pd, lsd, usd)), sep = '_'),
    age       = p0:(p0 + pd + lsd + usd - 1))
}

expand_completion_ages <- function(p0, pd, lsd, usd, offset = 3:5) {
  tibble(
    indicator = paste('comp', rep(c('prim', 'lsec', 'usec'), each = 3), 'v2', sep = '_'),
    age       = as.vector(outer(offset, cumsum(c(p0 + pd - 1, lsd, usd)), "+")))
}

expand_literacy_ages <- function(...) {
  tibble(
    indicator = 'literacy_1524',
    age       = 15:24)
}

indicator_age_scope <- 
  bind_rows(
    expand_ages(school_structure, expand_attendance_ages),
    expand_ages(school_structure, expand_completion_ages),
    expand_ages(school_structure, expand_literacy_ages)
    )

print(indicator_age_scope)
```

Remaining to be done:

- [ ] test and fine-tune articulation with WIDE microdata format, i.e. exact indicator names, correct age variable etc.
- [ ] add `expand_*_ages` function for remaining indicators

